\documentclass[10pt, a4paper]{amsart}
\input{preamble}

\usepackage{geometry}

\def\phi{\varphi}

\linespread{1.1}

\title{Seminar on Pila--Wilkie point counting and applications}
\author{Reid Barton \and Johan Commelin}

\begin{document}

\maketitle

\begin{abstract}
  \textbf{Keywords:}
  o-minimality, diophantine geometry, detailed proofs
  
  \medskip\noindent
  \textbf{Tl;dr:}
  In this seminar,
  we will refine our understanding of o-minimal theory
  by studying in detail the proof of the Pila--Wilkie point counting theorem.
  After that, we will look at applications of this result in diophantine geometry.

  \medskip\noindent
  \textbf{Main reference:}
  We will follow a survey article by Scanlon~\cite{scanlon}.
\end{abstract}

\section*{Appetizer}
Consider the following statement (a special case of Manin--Mumford):

\begin{theorem}
  Let $n > 0$ be a natural number,
  and let $\mathbb G = (\CC^*)^n$
  be the $n$-th power of the unit group of the complex numbers.
  Let $f \in \CC[x_1, \dots, x_n]$ be a polynomial in $n$ variables.
  Then the set
  \[
    X = \{ (\zeta_1, \dots, \zeta_n) \in \mathbb G \mid
    \text{each $\zeta_i$ is a root of unity and $f(\zeta_1, \dots, \zeta_n) = 0$}\}
  \]
  is a finite union of cosets of subgroups of $\mathbb G$.
\end{theorem}

Originally, this statement was proven by Mann,
but we will be interested in the Pila--Zannier argument.
It goes as follows.
Let $\exp \colon \CC \to \CC^*$ denote the function $z \mapsto e^{2\pi iz}$.
Observe that there is an analytic covering $\exp \colon \CC^n \to \mathbb G$.
A tuple $\zeta = (\zeta_1, \dots, \zeta_n)$ consists of roots of unity
if and only if there is some rational $a \in \QQ^n$ such that $\exp(a) = \zeta$.
This means that we can translate our problem into
a question about rational solutions to the transcendental equation $f(\exp(z)) = 0$.

It may seem as if we have made the problem a lot more difficult.
However, by restricting to a fundamental domain
\[
  D = \{ (z_1, \dots, z_n) \in \CC^n \mid 0 \le \Re(z_i) < 1 \text{ for each $i$}\}
\]
we end up in a \emph{tame} situation.

\begin{enumerate}
  \item It is again the case that $\zeta \in \mathbb G$ is a tuple of roots of unity
    if and only if there exists a \emph{rational} $a \in D \cap \QQ^n$ such that $\exp(a) = \zeta$.
  \item The restriction of~$\exp$ to~$D$ is a definable function in the structure~$\RR_{\an,\exp}$,
    which is an \emph{o-minimal} structure.
    From the point of view of mathematical logic, this means it is exceedingly well-behaved.
\end{enumerate}

We may now consider the set
\[ \tilde X = \{ z \in D \mid f(\exp(z)) = 0 \} \]
which is an example of a \emph{definable set} in~$\RR_{\an,\exp}$.
This is where the Pila--Wilkie point counting theorem comes in.

Understanding the rational solutions to algebraic equations is a notoriously hard problem.
But it turns out that one can get a good grip on rational solutions to \emph{transcendental} equations.

Let $Y \subset \RR^m$ be any set, definable in some o-minimal expansion of the reals.
We define the \emph{algebraic part} $Y^\alg \subset Y$ to be
the union of all connected, positive dimensional semialgebraic subsets of~$Y$.
(Recall that a set is semialgebraic if it is definable using Boolean combinations of polynomial inequalities.)
Next, we define the \emph{transcendental part} of~$Y$ to be $Y - Y^\alg$.

The Pila--Wilkie point counting theorem asserts that there are sub-exponentially many rational points in~$Y^\tra$.
To make this precise, we introduce the following function,
which counts rational points of \emph{bounded height}:
\[
  N(Y,t) = \# \{ (\tfrac{a_1}{b_1}, \dots, \tfrac{a_n}{b_n}) \in Y^\tra \mid
  \text{for each $i$ we have }
  |a_i| \le t, |b_i| \le t, a_i, b_i \in \ZZ \}
\]

\begin{theorem}[Pila--Wilkie]
  For each $\epsilon > 0$ there is a constant $C = C_\epsilon$
  so that $N(Y, t) \le Ct^\epsilon$ for all $t \ge 1$.
\end{theorem}

Now we return to our definable set of interest: $\tilde X$.
Using a result by Ax (a function field version of the Schanuel conjecture)
we can show that $\tilde X^\alg$ is indeed a finite union of cosets of subgroups
(intersected with $D$).

Hence we are done if we show that $\tilde X^\tra$ only has finitely many rational points.
This is done by contradiction.
If there are infinitely many points,
then one can use Galois theory
to show that $N(Y, t)$ must exhibit exponential growth.
This contradicts Pila--Wilkie, so we win.

\section{Introduction}

Give an overview of the seminar.
A long form of the appetizer above,
with a bit more details on what is meant with an o-minimal structure.

Besides Manin--Mumford for tori,
mention applications to Manin--Mumford for abelian varieties
as well as the Andr\'e--Oort conjecture for Shimura varieties.

\subsection*{Reference} \S1 and \S2 (aka p1--5) of~\cite{scanlon}

\section{O-minimality I: crash course on logic}

Recap: language, structure, definable set, formula, sentence, theory, model.

In particular, recap the notation $\mathfrak M \models \phi$.

State compactness! Explain what it means!

\subsection*{Reference} \S3.1 (aka p6--10) of~\cite{scanlon}

\section{O-minimality II: examples}

State theorems of Tarski and Wilkie.
Discuss proof of Tarski's theorem in some detail?

Explain what it means that $\RR_{\exp}$ is model complete,
but don't discuss the proof of Wilkie's theorem.

State theorem by Van den Dries, that $\RR_\an$ is o-minimal.

Finally, discuss $\RR_{\an,\exp}$.

\subsection*{Reference} first half of \S3.2 (aka p10--14) of~\cite{scanlon}

\section{O-minimality III: cell decomposition}

Prove the existence of Skolem functions.

Define cells, state cell decomposition. Discuss the proof.
(Note: a detailed proof is long and complicated.)

Derive Lemma~3.35 of~\cite{scanlon} on non-constant definable analytic curves
in definable sets.

\subsection*{Reference} second half of \S3.2 (aka p14--18) of~\cite{scanlon}

\section{Pila--Wilkie I: overview and structure of proof}

Define the multiplicative height of rational numbers and the point counting function.

Define the algebraic part and the transcendental part of a definable set.

State the Pila--Wilkie theorem.

State the two main ingredients in the proof:
Theorem~4.8 and Theorem~4.31 (from \S4.3)

State and prove Proposition~4.11 (uses compactness, take your time) and Proposition~4.15.

\subsection*{Reference} \S4.1 of~\cite{scanlon}.

\section{Pila--Wilkie II: parametrization theorem}

The goal is to prove Theorem 4.8.

State Theorem 4.19. Carefully outline the induction strategy.

Carry out the proof.

\subsection*{Reference} Last bit of \S4.1 and all of \S4.2 of~\cite{scanlon}.

\section{Pila--Wilkie III: diophantine approximation}

The goal is to prove Theorem~4.31.

Introduce notations,
and state and prove Propositions~4.28 and~4.30.

This section includes a lot of complicated formulas.
Think carefully about the best way to present this material.

\subsection*{Reference} \S4.3 of~\cite{scanlon}.

\section{Pila--Wilkie IV: end of proof, refinements}

Recap the proof structure:
recall statement of Pila--Wilkie, Proposition~4.11, and Theorem~4.31.

Tie everything together: State Proposition~4.33

State Theorem~4.32 (generalization of Pila--Wilkie) and prove it.

Talk about refinements.

\subsection*{Reference} \S4.4 of~\cite{scanlon}.

\section{Applications I: Manin--Mumford for tori}

Make the appetizer precise.

\subsection*{Reference}
See \S5 of~\cite{csp} for detailed calculations.

\section{Applications II: Manin--Mumford for abelian varieties}

The goal is to state and prove Theorem~5.1.

Along the way state and assume Theorems~5.5 and~5.7.

\subsection*{Reference} \S5.1 of~\cite{scanlon}.

\printbibliography

\end{document}

% vim: ts=2 et sw=2 sts=2

